Variance reduction for MCMC methods via martingale representations

### 2020-03-19 aap/Revision/vr-mcmc-revision-new-new.tex
update

\def\rmd{\mathrm{d}}
\def\rme{\mathrm{e}}
\def\rset{\mathbb{R}}
\def\NtrainPath{T}
\def\TrainSet{\mathcal{T}}
\modulolinenumbers[5]
\newcommand{\proofendsign}{$\Box$}
\newtheorem{thm}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{prop}[thm]{Proposition}
\newtheorem{remark}{Remark}
\newtheorem{cor}{Corollary}
\newtheorem{example}{Example}
\newenvironment{proof}{{\noindent \bf Proof }}
 {{\hspace*{\fill}\proofendsign\par\bigskip}}
\def\TV{\operatorname{TV}}
\newcommand{\jac}[1]{\operatorname{J} _{#1}}

@@

The situation becomes much more difficult in the case of MCMC algorithms, where one  has to work with a
Markov chain \(X_p,\) \(p=0,1,2,\ldots,\) whose marginal distribution  converges  to \(\pi\) as time grows. One important class of the variance reduction methods in this case  is based on the Poisson equation for the corresponding  Markov chain. It was observed in~\cite{henderson1997variance}  that if a time-homogeneous Markov chain \((X_p)\) is stationary with stationary distribution \(\pi,\) then for any real-valued function \(G \in L^1(\pi) \) defined on the state space of the Markov chain \((X_p),\)  the function \(U(x) := G(x)-\mathsf{E}[G(X_{1})|X_0 = x]\) has zero mean with respect to \(\pi\).  The best choice for the function \(G\) corresponds to a solution of the Poisson equation  \(\mathsf{E}[G(X_{1})|X_0 = x]-G(x)=-f(x)+\pi(f)\).  Moreover, it is also related to the minimal asymptotic variance in the corresponding central limit theorem, see \cite{duncan2016variance} and \cite{mira2013zero}.   Although the Poisson equation involves the quantity of interest \(\pi(f)\)  and can not be  solved explicitly in most cases, this idea still can be used to construct some  approximations for the optimal zero-variance control variates. For example,  \cite{henderson1997variance} proposed to compute approximations for the solution of the Poisson equation for specific Markov chains with particular emphasis on models arising in stochastic network theory. In \cite{dellaportas2012control} and \cite{brosse2018diffusion}  series-type control variates are introduced and studied for reversible Markov chains. It is assumed in \cite{dellaportas2012control}  that the one-step conditional expectations  can be computed explicitly  for a set of basis functions. The authors in \cite{brosse2018diffusion} proposed another approach tailored to diffusion setting which does not require the computation of integrals of basis functions and only involves  applications of the underlying generator.
\par
In this paper we focus on the  Langevin type algorithms which got much attention recently, see \cite{dalalyan2017theoretical,durmus:moulines:2017, MR2353037, MR3861816, MR2977986} and references therein. We propose  a generic variance reduction method for these and other types algorithms, which is purely non-asymptotic and does not require that    the conditional expectations of the corresponding Markov chain can be computed  or that the generator is known analytically. Moreover, we do not need to assume stationarity or/and sampling under the invariant distribution \(\pi.\) We rigorously analyse the convergence of the method and study its complexity. It is shown that  our variance-reduced Langevin algorithm outperforms the standard Langevin algorithms in terms of complexity.
\par
The paper is organized as follows.  In Section~\ref{sec:setup} we set up the problem and introduce some notations. Section~\ref{seq:mart_repr} contains a novel martingale representation and shows how this representation can be used for variance reduction. In Section~\ref{sec:ula_analysis} we analyze the performance of the proposed variance reduction algorithm in the case of Unadjusted Langevin Algorithm (ULA). Section~\ref{sec:coeff} studies the complexity of the variance reduced ULA. Finally, numerical examples are presented in Section~\ref{sec:num}.

@@

\section{Martingale representation and variance reduction}

@@

$\phi_k(\xi)$, $k\ge1$, are centered. As an example, we can take  multivariate Hermite polynomials for the ULA algorithm and a tensor product of Shifted Legendre polynomials for "uniform part" and Hermite polynomials for "Gaussian part"  of the random variable $\xi = (u, z)^T$ in MALA, as the Shifted Legendre polynomials are orthogonal with respect to the Lebesgue measure on \([0,1].\)  Let $\left(\Omega,\mathcal{F},\mathsf{P}\right)$ be a probability space, $(\xi_p)_{p \in \nset}$ be i.i.d. $m-$dimensional random vectors. We denote by $(\mathcal{G}_p)_{p \in \nset_0}$  the filtration generated by $(\xi_p)_{p \in \nset}$ with the convention $\mathcal{G}_0=\mathrm{triv}$. For  $x \in \rset^d, y \in \rset^{m}$ and $k \in \nset$ let $\Phi_k(x,y)$ be a function mapping $\rset^{d+m}$ to $\rset^d$. Then we can define for $l \le p$

@@

X^x_{l,p}=G_{l,p}(x,\xi_{l},\ldots,\xi_{p}),

@@

with the function \(G_{l,p}:\) \(\rset^{d + d_{\xi}\times(p-l+1)}\to \rset^{d}\) defined as

@@

G_{l,p}(x,y_l,\ldots,y_p):=\Phi_p(\cdot,y_{p})\circ\Phi_{p-1}(\cdot,y_{p-1})\circ\dots\circ\Phi_{l}(x,y_{l})

@@

Note that $\left(X^x_{0,p}\right)_{p \in \nset_0}$ is a Markov chain with values in $\rset^d$ of the form \eqref{eq:chain_gen}, starting at $X_0 = x$. We write $X^x_{p}$ and $G_{p}$ as a shorthand notation for $X^x_{0,p}$ and $G_{0,p}$, respectively.

\begin{thm}\label{prop:29032018a1}
For all $p \in \nset,$ $q \leq p$, $j < q \leq p$, Borel bounded functions $f: \rset^d \rightarrow \rset$ and $x \in \rset^d$ the following representation holds in \(L^2(\mathsf{P})\)
\begin{eqnarray}
\label{eq:mart_repr}
f(X^x_{q})=\mathsf{E}\left[\left.f(X^x_{q})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{q}a_{q,l,k}(X^x_{l-1})\phi_k\left(\xi_{l}\right),
\end{eqnarray}
where \((X^x_{p})_{p\geq 0}\) is given in \eqref{eq:ula_new} and for all $y \in \rset^d$
\begin{eqnarray}
\label{eq:coeff_mart}
%a_{p,l,k}(x)=\mathsf{E}\left[\left.f(X_{p})\phi_k\left(\xi_{l}\right)\right|X_{l-1}=x\right], \quad p\geq l, \quad k\in \mathbb{N}.
a_{q,l,k}(y)=\mathsf{E}\left[f(X^y_{l-1,q})\phi_k\left(\xi_{l}\right)\right], \quad q\geq l, \quad k\in \nset
\end{eqnarray}
\end{thm}

\begin{proof}%[Proof of Theorem~\ref{prop:29032018a1}]
The expansion obviously holds for $p = q=1$ and $j=0$.
Indeed, since $\left(\phi_{k}\right)_{k \geq 0}$ is a complete orthonormal system in \(L^2(\mathbb{R}^d, P_{\xi})\), it holds in \(L^2(\mathsf{P})\) that
\[
f(X^x_{1})=\PE[f(X^x_{1})]+\sum_{k\geq1}a_{1,1,k}(x)\phi_{k}(\xi_{1})
\]
for any bounded $f$ with $a_{1,1,k}(x)=\PE[f(X_{1}^x)\phi_{k}(\xi_{1})]$.
Assume now that (\ref{eq:mart_repr}) holds for some $m < p$, any $q \leq m$, $j < q \leq m$ and all bounded $f$. Let us prove that the induction assumption holds for $q=m+1$ and all $j < q$.
The orthonormality and completeness
of the system $\left(\phi_{k}\right)_{k=0}^\infty$ implies that for any bounded $f$ and $y \in \rset^d$, 
$\lim_{n \rightarrow \infty}\Psi_{n,m+1,1}(y)= 0$ with $\Psi_{n,m+1,1}(y)= \PE[|f(X^y_{m,m+1}) - f_{n,m+1,1}(y)|^2]$ and
\[
f_{n,m+1,1}(y) = \PE[f(X^y_{m,m+1})] +  \sum\limits_{k=1}^{n}\PE[f(X^y_{m,m+1})\phi_{k}(\xi_{m+1})]\phi_k(\xi_{m+1}).
\]
Note that  for any $y \in \rset^d$ and $n \in \nset$ it holds $\Psi_{n,m+1,1}(y) \leq \Psi_{0,m+1,1}(y)$ and
\begin{eqnarray*}
\PE \left[ \psi_{0,m+1,1}(X_m^x) \right] &=& \PE \left[ \left. \PE \left[ \psi_{0,m+1,1}(X_m^x) \right| \mathcal{G}_m\right]\right] = \PE \left[ \left. \PE \left[ |f(X^{X_m^x}_{m,m+1}) - \PE f(X^{X_m^x}_{m,m+1})|^2 \right| X_m^x\right]\right]
\\
&=& \PE \left[ |f(X_{m+1}^x) - \PE f(X_{m+1}^x)|^2 \right] = \PVar{f(X_{m+1}^x)} < \infty
\end{eqnarray*}
Hence, by Lebesgue dominated convergence theorem, $\lim_{n \rightarrow \infty}\PE \left[\psi_{n,m+1,1}(X_m^x)\right] = 0$ and since for all $y \in \rset^d$ the expectation $\PE[f(X^y_{m,m+1})]$ is a version of $\PE \left[f(X^y_{m,m+1}) | \mathcal{G}_m \right]$, it holds in \(L^2(\mathsf{P})\) that
\begin{equation}
\label{eq:1-step-decomp}
f(X^x_{m+1}) = \PE[f(X^x_{m+1})|\mathcal{G}_m] + \sum\limits_{k=1}^{\infty}a_{m+1,m+1,k}(X^x_{m})\phi_{k}(\xi_{m+1})
\end{equation}
where
\[
a_{m+1,m+1,k}(y) = \PE \left[f(X^y_{m,m+1})\phi_k\left(\xi_{m+1}\right)\right]
\]
which is the required statement in the case $q = m+1$ and $j=m$.
Now assume that $j<m$.
Set $g(y) = \PE \left[f(X_{m,m+1}^y)\right]$. Note that $\mathsf{P}$-a.s. it holds $g(X_m^x) = \PE\left[\left.f(X_m^x) \right| \mathcal{G}_m\right]$ and $g$ is bounded by construction. Hence the induction hypothesis applies, and we get
\begin{equation}\label{eq:28082017a1}
\mathsf{E}\left[\left.f(X^x_{m+1})\right| \mathcal{G}_m \right]=\mathsf{E}\left[\left.f(X^x_{m+1})\right| \mathcal{G}_j \right]+\sum_{k\geq1}\sum_{l=j+1}^{m}a_{m+1,l,k}(X^x_{l-1})\phi_{k}(\xi_{l})
\end{equation}
with
\begin{align*}
a_{m+1,l,k}(X^x_{l-1}) &= \mathsf{E}\left[\left.\mathsf{E}\left[\left.f(X^x_{m+1})\right|\mathcal{G}_{m}\right]\phi_{k}(\xi_{l})\right|\mathcal{G}_{l-1}\right]
 = \mathsf{E}\left[\left.f(X^x_{m+1})\phi_{k}(\xi_{l})\right|X^x_{l-1}\right].
\end{align*}
where for $y \in \rset^d$,
\[
a_{m+1,l,k}(y) = \mathsf{E}\left[f(X^y_{l-1,m+1})\phi_{k}(\xi_{l})\right]
\]
Formulas \eqref{eq:1-step-decomp}
and~\eqref{eq:28082017a1} conclude the induction step for $q = m+1$ and all $j < q$ and hence the proof.
\end{proof}

\begin{remark}\label{rem:general_moments} Denote by $\mathcal{F}_2^p$ class of Borel functions $f: \rset^d \to \rset$ such that for any $n \leq p$, any indices $1 \leq i_1 < \ldots < i_n \leq p$ and $y \in \rset^d$,
\begin{equation}
\label{def:f_2_p}
\mathsf{E}\left[\left |f(\Phi_{i_n}(\cdot,\xi_{i_n})\circ\Phi_{i_{n-1}}(\cdot,\xi_{i_{n-1}})\circ\dots\circ\Phi_{i_{1}}(y,\xi_{i_{1}}))\right|^{2}\right]<\infty
\end{equation}
Then the statement of Theorem~\ref{prop:29032018a1} remains valid for $f \in \mathcal{F}_2^p$. 
\end{remark}

@@

\begin{cor}
Let \((X^x_{p})_{p\geq 0}\) be a homogeneous Markov chain of form \eqref{eq:ula_new} with $\Phi_l = \Phi,$ $l\geq 1$. Then for all $p \in \nset,$ $q \leq p$, $j < q \leq p$, $f \in \mathcal{F}_{2,\text{hom}}^p$ and $x \in \rset^d$ it holds in \(L^2(\mathsf{P})\)
\begin{eqnarray*}
f(X^x_{q})=\mathsf{E}\left[\left.f(X^x_{q})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{q}\bar a_{q-l,k}(X^x_{l-1})\phi_k\left(\xi_{l}\right)
\end{eqnarray*}
where for $y \in \rset^d$
\begin{eqnarray*}
\bar a_{r,k}(y)=\mathsf{E}\left[f(X^y_{r})\phi_k\left(\xi_{1}\right)\right], \quad r, k\in \mathbb{N}.
\end{eqnarray*}
\end{cor}

From  numerical point of view another representation of the coefficients \(a_{p,l,k}\)  turns out to be more useful.
\begin{prop}
Let $q\geq l, k\in \nset$. Then the coefficients \(a_{q,l,k}(x)\) in \eqref{eq:coeff_mart}  can be alternatively represented as
\begin{eqnarray*}
a_{q,l,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{q,l}\left(\Phi_l(x,\xi)\right)\right]
\end{eqnarray*}
with \(Q_{q,l}(x)=\mathsf{E}\left[f(X^x_{l,q})\right],\) \(q\geq l.\)
%where   a r.v. \(\xi\) independent of \(\mathcal{G}_{l-1}.\)
The functions \((Q_{q,l})_{l=0}^q\)  can be computed by the backward recurrence: $Q_{q,q}(x)=f(x)$ and for $l \in \{0,\ldots,q-1\}$
\begin{eqnarray}
\label{eq:qpl}
Q_{q,l}(x)=\mathsf{E}\left[Q_{q,l+1}(X^x_{l+1})\right].
%=\mathsf{E}[Q_{p,l+1}(\Phi_{l+1}(x,\xi))]
\end{eqnarray}
In the case $\Phi_l=\Phi$ in \eqref{eq:ula_new} for all $l\geq 1,$ we have
\begin{equation}
\label{eq:definition-bar-a}
\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right]
\end{equation}
with  \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) $r\in \mathbb{N}.$
\end{prop}
\par
Next we show how  the representation \eqref{eq:mart_repr} can be used to  reduce the variance of MCMC algorithms.  Consider the time homogeneous case ($\Phi_l = \Phi$ for all $l \in \nset$) and the sample average
\begin{equation}\label{eq:29032018a2}
\pi_{n}^{N}(f)=\frac{1}{n}\sum_{p=N+1}^{N+n} f(X^x_{p}),
\end{equation}
where $N\in\mathbb N_0$ is the length of the burn-in period and $n\in\mathbb N$
the number of effective samples.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
\begin{eqnarray}
\nonumber
M_{K,n}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{k=1}^{K}\sum_{l=N+1}^{p} \bar a_{p-l,k}(X^x_{l-1})\phi_k(\xi_{l})\right]
\\
&=& \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\sum_{p=0}^{N+n-l}\bar{a}_{p,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].
\label{eq:29032018a5}
\end{eqnarray}
Since \(X^x_{l-1}\) is independent of \(\xi_{l}\) and \(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq 0\), we obtain
\[
\PE[g(X^x_{l-1})\phi_{k}\left(\xi_{l}\right)] = \mathsf{E}[g(X^x_{l-1})\CPE{\phi_{k}\left(\xi_{l}\right)}{\mathcal{G}_{l-1}]} = 0
\]
for any measurable function $g$  and hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
The representation \eqref{eq:coeff_mart} suggests that the variance of  the  variance-reduced estimator
\begin{equation}
\label{eq:29032018a3}
\pi_{K,n}^N(f)=\pi_n^N(f)-M_{K,n}^N(f)
\end{equation}
should be small for \(K\) large enough. Indeed, since $\PE[ \phi_k(\xi_l) \phi_{k'}(\xi_l)]=0$ if $k \ne k'$, we obtain
\begin{equation}
\label{eq:expression-variance}
\PVar[\pi_{K,n}^N(f)]=\frac{1}{n^2}\sum_{k=K+1}^{\infty}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X^x_{N+l-1})],
\end{equation}
where 
\begin{eqnarray}
\label{eq:definition-A-s,k}
A_{q,k}(y)=\sum_{r=0}^q \bar{a}_{r,k}(y),\quad q\in \mathbb{N},
\end{eqnarray}
and $\mathsf{Var}[\pi_{K,n}^N(f)]$ is small provided that the coefficients $A_{s,k}$ decay fast enough as $k\to \infty.$
In the next section   we provide a detailed   analysis of  $\mathsf{Var}[\pi_{K,n}^N(f)]$  for the  ULA  algorithm (see Example~\ref{exam:langevin-algorithm}). In many situations the coefficients $\bar{a}_{r,k}$ also decay exponentially fast as $r$ increases. The following result can be easily established.
\begin{lem}
\label{lem:coef_decrease}
Let $(X^x_{p})_{p \in \nset_0}$ be a $V-$geometrically ergodic homogeneous Markov chain with stationary distribution $\pi$ and $f$ be a Borel function with $\|f\|_{V} < \infty$. Then it holds
\begin{equation}
|\bar a_{r,k}(x)| \leq C\rho^r\|f\|_{V} (\PE V^2(\Phi(x,\xi)))^{1/2}
\end{equation}
where $\rho\in (0,1)$ and $C$ is an absolute constant. In particular, if $f$ is uniformly bounded we can take $V(x)\equiv 1$ to get
\begin{eqnarray*}
|\bar a_{r,k}(x)| \leq C\rho^r\|f\|_{\infty}.
\end{eqnarray*}

\end{lem}
This lemma shows that in fact  only a logarithmic number (in $n$) of coefficients need to be computed.

>>>>

\def\rmd{\mathrm{d}}
\def\rme{\mathrm{e}}
\def\rset{\mathbb{R}}
\def\NtrainPath{T}
\def\TrainSet{\mathcal{D}}
\modulolinenumbers[5]
\newcommand{\proofendsign}{$\Box$}
\newtheorem{thm}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{prop}[thm]{Proposition}
\newtheorem{remark}{Remark}
\newtheorem{cor}{Corollary}
\newtheorem{example}{Example}
\newenvironment{proof}{{\noindent \bf Proof }}
 {{\hspace*{\fill}\proofendsign\par\bigskip}}
\def\TV{\operatorname{TV}}
\newcommand{\jac}[1]{\operatorname{J} _{#1}}

@@

The situation becomes much more difficult in the case of MCMC algorithms, where one  has to work with a
Markov chain \(X_p,\) \(p=0,1,2,\ldots,\) whose marginal distribution  converges  to \(\pi\) as time grows. One important class of the variance reduction methods in this case  is based on the Poisson equation for the corresponding  Markov chain. It was observed in~\cite{henderson1997variance}  that if a time-homogeneous Markov chain \((X_p)\) is stationary with stationary distribution \(\pi,\) then for any real-valued function \(G \in L^1(\pi) \) defined on the state space of the Markov chain \((X_p),\)  the function \(U(x) := G(x)-\mathsf{E}[G(X_{1})|X_0 = x]\) has zero mean with respect to \(\pi\).  The best choice for the function \(G\) corresponds to a solution of the Poisson equation  \(\mathsf{E}[G(X_{1})|X_0 = x]-G(x)=-f(x)+\pi(f)\).  Moreover, it is also related to the minimal asymptotic variance in the corresponding central limit theorem, see \cite{duncan2016variance} and \cite{mira2013zero}.   Although the Poisson equation involves the quantity of interest \(\pi(f)\)  and can not be  solved explicitly in most cases, this idea still can be used to construct some  approximations for the optimal zero-variance control variates. For example,  \cite{henderson1997variance} proposed to compute approximations for the solution of the Poisson equation for specific Markov chains with particular emphasis on models arising in stochastic network theory. In \cite{dellaportas2012control} and \cite{brosse2018diffusion}  series-type control variates are introduced and studied for reversible Markov chains. It is assumed in \cite{dellaportas2012control}  that the one-step conditional expectations  can be computed explicitly  for a set of basis functions. The authors in \cite{brosse2018diffusion} proposed another approach tailored to diffusion setting which does not require the computation of integrals of basis functions and only involves  applications of the underlying generator.
\par
In this paper we focus on the  Langevin type algorithms which got much attention recently, see \cite{dalalyan2017theoretical,durmus:moulines:2017, MR2353037, MR3861816, MR2977986} and references therein. We propose  a generic variance reduction method for these and other types algorithms, which is purely non-asymptotic and does not require that    the conditional expectations of the corresponding Markov chain can be computed  or that the generator is known analytically. Moreover, we do not need to assume stationarity or/and sampling under the invariant distribution \(\pi.\) We rigorously analyse the convergence of the method and study its complexity. It is shown that  our variance-reduced Langevin algorithm outperforms the standard Langevin algorithms in terms of complexity.
\par
The paper is organized as follows.  In Section~\ref{sec:setup} we set up the problem and introduce some notations. Section~\ref{seq:mart_repr} contains a novel martingale representation and shows how this representation can be used for variance reduction. In Section~\ref{sec:ula_analysis} we analyze the performance of the proposed variance reduction algorithm in the case of Unadjusted Langevin Algorithm (ULA).  Finally, numerical examples are presented in Section~\ref{sec:num}.

@@

\section{Martingale representation}

@@

$\phi_k(\xi)$, $k\ge1$, are centered. As an example, we can take  multivariate Hermite polynomials for the ULA algorithm and a tensor product of shifted Legendre polynomials for "uniform part" and Hermite polynomials for "Gaussian part"  of the random variable $\xi = (u, z)^T$ in MALA, as the shifted Legendre polynomials are orthogonal with respect to the Lebesgue measure on \([0,1].\)  
Let  $(\xi_p)_{p \in \nset}$ be i.i.d. $m-$dimensional random vectors. We denote by $(\mathcal{G}_p)_{p \in \nset_0}$  the filtration generated by $(\xi_p)_{p \in \nset}$ with the convention $\mathcal{G}_0=\mathrm{triv}$. For  $x \in \rset^d, y \in \rset^{m}$ and $k \in \nset,$ let $\Phi_k(x,y)$ be a function mapping $\rset^{d+m}$ to $\rset^d$. Then we set for $l \le p$

@@

X^x_{l,p}:=G_{l,p}(x,\xi_{l},\ldots,\xi_{p}),

@@

with the functions \(G_{l,p}:\) \(\rset^{d + m\times(p-l+1)}\to \rset^{d}\) defined as

@@

G_{l,p}(x,y_l,\ldots,y_p):=\Phi_p(\cdot,y_{p})\circ\Phi_{p-1}(\cdot,y_{p-1})\circ\dots\circ\Phi_{l}(x,y_{l}).

@@

Note that $\left(X^x_{0,p}\right)_{p \in \nset_0}$ is a Markov chain with values in $\rset^d$ of the form \eqref{eq:chain_gen}, starting at $X_0 = x$. In the sequel we write $X^x_{p}$ and $G_{p}$ as a shorthand notation for $X^x_{0,p}$ and $G_{0,p}$, respectively.

\begin{thm}\label{thm:main-repr}
For all $p \in \nset,$ $q \leq p$, $j < q \leq p$, any Borel bounded functions $f: \rset^d \rightarrow \rset$ and all $x \in \rset^d$ the following representation holds in \(L^2(\mathsf{P})\)
\begin{eqnarray}
\label{eq:mart_repr}
f(X^x_{q})=\mathsf{E}\left[\left.f(X^x_{q})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{q}a_{q,l,k}(X^x_{l-1})\phi_k\left(\xi_{l}\right),
\end{eqnarray}
where \((X^x_{p})_{p\geq 0}\) is given in \eqref{eq:ula_new} and for all $y \in \rset^d$
\begin{eqnarray}
\label{eq:coeff_mart}
%a_{p,l,k}(x)=\mathsf{E}\left[\left.f(X_{p})\phi_k\left(\xi_{l}\right)\right|X_{l-1}=x\right], \quad p\geq l, \quad k\in \mathbb{N}.
a_{q,l,k}(y)=\mathsf{E}\left[f(X^y_{l-1,q})\phi_k\left(\xi_{l}\right)\right], \quad q\geq l, \quad k\in \nset.
\end{eqnarray}
\end{thm}

\begin{remark}\label{rem:general_moments} Denote by $\mathcal{F}_2^p$ class of Borel functions $f: \rset^d \to \rset$ such that for any $n \leq p$, any indices $1 \leq i_1 < \ldots < i_n \leq p$ and $y \in \rset^d$,
\begin{equation}
\label{def:f_2_p}
\mathsf{E}\left[\left |f(\Phi_{i_n}(\cdot,\xi_{i_n})\circ\Phi_{i_{n-1}}(\cdot,\xi_{i_{n-1}})\circ\dots\circ\Phi_{i_{1}}(y,\xi_{i_{1}}))\right|^{2}\right]<\infty
\end{equation}
Then the statement of Theorem~\ref{thm:main-repr} remains valid for $f \in \mathcal{F}_2^p$. 
\end{remark}

@@

\begin{cor}
Let \((X^x_{p})_{p\geq 0}\) be a homogeneous Markov chain of the form \eqref{eq:ula_new} with $\Phi_l = \Phi,$ $l\geq 1$. Then for all $p \in \nset,$ $q \leq p$, $j < q \leq p$, $f \in \mathcal{F}_{2,\text{hom}}^p$ and $x \in \rset^d$ it holds in \(L^2(\mathsf{P})\)
\begin{eqnarray*}
f(X^x_{q})=\mathsf{E}\left[\left.f(X^x_{q})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{q}\bar a_{q-l,k}(X^x_{l-1})\phi_k\left(\xi_{l}\right)
\end{eqnarray*}
where for all $y \in \rset^d$
\begin{eqnarray*}
\bar a_{r,k}(y)=\mathsf{E}\left[f(X^y_{r})\phi_k\left(\xi_{1}\right)\right], \quad r, k\in \mathbb{N}.
\end{eqnarray*}
\end{cor}

Next we show how  the representation \eqref{eq:mart_repr} can be used to  reduce the variance of MCMC algorithms.  For the sake of clarity, in the sequel we consider only the time homogeneous case ($\Phi_l = \Phi$ for all $l \in \nset$). Define
\begin{equation}\label{eq:29032018a2}
\pi_{n}^{N}(f)=\frac{1}{n}\sum_{p=N+1}^{N+n} f(X^x_{p}),
where $N\in\mathbb N_0$ is the length of the burn-in period and $n\in\mathbb N$
the number of effective samples.
Fix some $K\in\mathbb N$ and denote
M_{K,n}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{k=1}^{K}\sum_{l=N+1}^{p} \bar a_{p-l,k}(X^x_{l-1})\phi_k(\xi_{l})\right]
&=& \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\sum_{p=0}^{N+n-l}\bar{a}_{p,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].
\label{eq:29032018a5}
Since \(X^x_{l-1}\) is independent of \(\xi_{l}\) and \(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq 0\), we obtain
\PE[g(X^x_{l-1})\phi_{k}\left(\xi_{l}\right)] = \mathsf{E}[g(X^x_{l-1})\CPE{\phi_{k}\left(\xi_{l}\right)}{\mathcal{G}_{l-1}]} = 0
for any measurable function $g.$  Hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
The representation \eqref{eq:coeff_mart} suggests that the variance of  the   estimator
\label{eq:29032018a3}
\pi_{K,n}^N(f)=\pi_n^N(f)-M_{K,n}^N(f)
should be small for \(K\) large enough. Indeed, since $\PE[ \phi_k(\xi_l) \phi_{k'}(\xi_l)]=0$ if $k \ne k'$, we obtain
\label{eq:expression-variance}
\PVar[\pi_{K,n}^N(f)]=\frac{1}{n^2}\sum_{k=K+1}^{\infty}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X^x_{N+l-1})],
\label{eq:definition-A-s,k}
A_{q,k}(y)=\sum_{r=0}^q \bar{a}_{r,k}(y),\quad q\in \mathbb{N}.
Hence $\mathsf{Var}[\pi_{K,n}^N(f)]$ is small provided that the coefficients $A_{s,k}$ decay fast enough as $k\to \infty.$
In   Section~\ref{sec:ula_analysis}   we provide a detailed  theoretical  analysis of  $\mathsf{Var}[\pi_{K,n}^N(f)]$  for the  ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).   
\section{Estimation of coefficients}
The coefficients \((\bar a_{l,k})\) need to be estimated before one can apply the proposed variance reduction approach. This amounts to solving a nonparametric regression problem.
We first present a generic regression algorithm and then in Section~\ref{sec:num} give further implementation details. Our algorithm starts with estimating the functions 
$A_{q,k}$ for $q=N+1,\ldots,N+n,$ $k=1,\ldots,K. $ We first generate \(\NtrainPath\)  paths  of the chain \(X\)
(the so-called ``training paths''):
\TrainSet=\Bigl\{(X^{(s)}_{N+1},\ldots,X^{(s)}_{N+n}),\quad s=1,\ldots,\NtrainPath\Bigr\}
with $X^{(s)}_{0}=x,$ $s=1,\ldots,T.$
Then we solve   the least squares optimization problems
\begin{equation}\label{eq:06042018a1}
\widetilde{A}_{q,k} \in \argmin_{\psi \in \Psi} \sum_{s = 1}^{T} \left| \sum_{r = 0}^{q}  f(X_{r+L}^{(s)})\phi_k(\xi^{(s)}_{L+1}) - \psi(X^{(s)}_L)\right|^2
for some $L$ with $L+q\leq N+n,$
where \(\Psi\) is a  class of  real-valued functions on \(\mathbb{R}^d\) and $\xi^{(s)}_{N+1},\ldots,\xi^{(s)}_{N+n}$ are iid random variables with distribution $P_\xi$ used to construct the $s$-th trajectory of $X.$   As usual in regression analysis (see Section~11 in \cite{gyorfi2006distribution})  we  truncate  the least squares estimates $(\widetilde{A}_{q,k})$ and set  
\label{eq:trunc}
\widehat{A}_{q,k}(y)=
\widetilde{A}_{q,k}(y), & |\widetilde{A}_{q,k}(y)|\leq U_k,
U_k \,\mathrm{sign}(\widetilde{A}_{q,k}(y)), & \mathrm{otherwise}. 
 for a sequence of positive real numbers $U_k,$ $k=1,\ldots,K.$
Upon estimating  the functions \((A_{q,k})\), one can construct the empirical estimate  of \(M_{K,n}^N(f)\) in the form
\widehat M_{K,n}^N(f) = \frac{1}{n}\sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left[\widehat{A}_{N+n-l,k}(X_{l-1}^{x})\phi_{k}(\xi_{l})\right].
Obviously, \(\mathsf{E}[\widehat M_{K,n}^N(f)|\TrainSet]=0\) and \(\widehat M_{K,n}^N(f)\) is a valid control variate in that it
does not introduce any bias. By the Jensen inequality and orthonormality of \((\phi_k)_{k\geq 0},\)
\begin{multline*}
\mathsf{E}\left[\left.\left|\widehat{M}_{K,n}^{N}(f)-M_{K,n}^{N}(f)\right|^{2}\right | \TrainSet \right]
\leq \frac{1}{n^2}\sum_{k=1}^{K}\sum_{l=1}^{n}\PE\left[\left.|A_{n-l,k}(X^x_{N+l-1})-\widehat A_{n-l,k}(X^x_{N+l-1})|^2\right | \TrainSet \right].
\end{multline*}
Set
\widehat\pi_{K,n}^N(f):=\pi_n^N(f)-\widehat{M}_{K,n}^N(f)
Combining this with \eqref{eq:expression-variance} results in 
\begin{multline*}
\PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]=\frac{1}{n^2}\sum_{k=K+1}^{\infty}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X^x_{N+l-1})]
+\frac{1}{n^2}\sum_{k=1}^{K}\sum_{l=1}^{n}\PE\left[\left.|A_{n-l,k}(X^x_{N+l-1})-\widehat A_{n-l,k}(X^x_{N+l-1})|^2\right | \TrainSet \right].
\end{multline*}
In the next section we analyze the case of ULA and show that under some rather mild conditions (convexity of the potential $U$ outside a ball around zero)
\PE[A^2_{l,k}(X^x_{N+l-1})]\leq R^2_k,\quad k=1,2,\ldots, 
for a sequence  $(R_k)$ satisfying $\sum_{k=1}^\infty R^2_k\leq C$ with constant $C$  depending neither on $\gamma$ nor on $n$ (see \eqref{eq:Aqk2}). Hence if we take $U_k=R_k$ as truncation parameters in \eqref{eq:trunc}, then $\widehat A_{n-l,k}$ converges to $A_{n-l,k}$ as $T\to \infty.$ Moreover,
\PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]\leq \frac{2C}{n}, 
for a constant $C$ not depending on $n$ and $\gamma.$
If the class of functions $\Psi$ is linear of dimension $\mathrm{D}_\Psi$ As can be easily seen the cost of computing $\widehat \pi_{K,n}^N(f)$ is of order $ K T^2 n$. At the same time the variance of the standard estimate $\widehat \pi_{n}^N(f)$ is of order $1/(n\gamma)$ and this bound can not be improved in general. Thus the ratio of cost-variances is of order
\frac{\mathrm{cost}(\pi_{K,n}^N) \PVar[\widehat \pi_{K,n}^N(f)|\TrainSet]}{\mathrm{cost}(\pi_{n}^N)\PVar[\pi_{n}^N(f)]}\leq C \gamma K T^2 L^2 \mathrm{Dim}(\Psi). 
